Iterative Monomorphisation

### 2025-05-12 iterative_monomorphisation.tex
shorten and thank

@@

\begin{sloppy}
\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose an iterative approach that is incomplete but pragmatic. We implemented it in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation demonstrates that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

@@

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed description to help future implementers. In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).

@@

@@

@@

The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It operates by applying a bounded number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase and an instantiation phase. Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.

@@

The algorithm presented above is too naïve for a practical implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop combinatorially explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This avoids having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are completed, in a separate, final step.

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. No new formulae are generated at this stage, only new type arguments and substitutions. Once these iterations are completed, the first phase is complete and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. This is when the new formulae are generated. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

Each bound of the monomorphisation process represents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time. Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

   We thank Sascha Böhme for fruitful discussions.
   We thank Jannis Limperg and Mark Summerfield for suggesting textual improvements.

>>>>

\hyphenation{mono-mor-phisa-tion}
\hyphenation{mono-mor-phic}
\hyphenation{non-mono-mor-phic}

@@

\begin{sloppy}
\begin{abstract}
Monomorphisation can be used to extend monomorphic provers to support polymorphic logics. We propose a pragmatic iterative approach. We implemented it in the Zipperposition prover, where it is used to translate away polymorphism before invoking the monomorphic prover E as a backend. Our evaluation shows that this approach increases Zipperposition's success rate. Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.

@@

In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}). We also provide a more detailed\pagebreak[2] \hbox{description} to help future implementers. In addition, our paper shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).

@@

\kern-1pt %%% TYPESETTING: Hack

@@

\pagebreak[2]

@@

The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem. It works by applying a bounded number of iterations, each taking a polymorphic problem as argument and returning a problem with new partially instantiated formulae. A single iteration consists of a collection phase and an instantiation phase. Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.

@@

\looseness=-1
The algorithm presented above is too naïve in practice. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop combinatorially explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This avoids having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are completed, in a separate, final step.

@@

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations,\pagebreak[2]
the old and new type argument maps are updated with newly generated types. No new formulae are generated at this stage, only new type arguments and substitutions. Once these iterations are completed, the first phase is complete and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. This is when the new formulae are generated. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

Each bound of the monomorphisation process represents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time. Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args}~to~\ref{e_settings}.

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-mono\-morphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

   We thank Sascha Böhme for fruitful discussions.
   We thank Jannis Limperg, Mark Summerfield, and three anonymous reviewers for suggesting textual improvements.
